---
title: "The Relationship between Anemia Prevalence in Young Women and Meat Consumption in Low & Middle Income Countries "
author: "Sylvia Baeyens"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---


```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
```


```{r, echo=FALSE}
# Initialize code chunk options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")

options(digits=2)
```

<br>

# Introduction

This PM566 final project will look at the relationship between meat consumption and anemia prevalence in young women in low and middle income countries. 

**This is a summary of the background, data visualization, and conclusion. The full report can be downloaded from the button on the top of this page or from the link at the bottom of this page**

Anemia is one of the most common medical conditions among menopausal women. Anemia is defined as a condition where the body lacks enough red blood cells to carry sufficient oxygen. Anemia is associated with many health risk factors and increased fatigue. 

The main cause of anemia is iron deficiency. Iron is required to produce hemoglobin, a protein in red blood cells which binds to oxygen. Another secondary cause of anemia is blood loss, which occurs cyclically, usually monthly, in menopausal women; therefore, menopausal women are at an increased risk for anemia. 

Iron can be consumed through diet in two forms: heme and non-heme iron. Heme iron is found only in animal products, such as red meat and poultry. Non-heme iron is plant based, and can be found in food sources such as nuts, legumes, and leafy greens. Heme iron is more readily absorbed by the body than non-heme iron; non-heme iron often requires secondary vitamins, such as vitamin C, to up-regulate proper absorption. While it is certainly possible to avoid an iron deficiency on a plant-based diet, most plant-based individuals choose to supplement their diet with iron supplements.

This project will focus solely on lower and middle income countries, as information on proper diet and supplementation is not readily available to these populations and the economic burden of anemia is greater.


```{r anemia data cleanup, echo= FALSE, include= FALSE}
#reading in all 3 anemia data sets: mild, moderate & severe
#changing mean variable to be representative of the type of anemia

#full url of data is http://ghdx.healthdata.org/record/ihme-data/global-anemia-prevalence-geospatial-estimates-2000-2019

mildAnemia= data.table::fread("data/MildAnemia.CSV") 
mildAnemia = mildAnemia %>%
  rename(COUNTRY = "ADM0_NAME", mildPct = "mean")

modAnemia= data.table::fread("data/ModerateAnemia.CSV") 
modAnemia = modAnemia %>%
  rename(COUNTRY = "ADM0_NAME", modPct = "mean")

sevAnemia= data.table::fread("data/SevereAnemia.CSV") 
sevAnemia = sevAnemia %>%
  rename(COUNTRY = "ADM0_NAME", sevPct = "mean")

# finding mean Anemia pct by country & year
mildAvg = mildAnemia[,.(
  mildPct = mean(mildPct,na.rm = TRUE)
  ), by = c("COUNTRY", "year")]
modAvg = modAnemia[,.(
  modPct = mean(modPct,na.rm = TRUE)
  ), by = c("COUNTRY", "year")]
sevAvg = sevAnemia[,.(
  sevPct = mean(sevPct,na.rm = TRUE)
  ), by = c("COUNTRY", "year")]

# merging all 3 anemia data frames
TotalAnemia = merge(
  x = mildAvg,
  y = modAvg,
  all.x = TRUE, all.y = FALSE
)

TotalAnemia = merge(
  x = TotalAnemia,
  y = sevAvg,
  all.x = TRUE, all.y = FALSE
)

#check for any missing data
TotalAnemia[,table(is.na(mildPct))]
TotalAnemia[,table(is.na(modPct))]
TotalAnemia[,table(is.na(sevPct))]
#no missing data
```


```{r meat data cleanup, echo= FALSE, include= FALSE}
#reading in meat dataset
# url : https://ourworldindata.org/meat-production#which-countries-eat-the-most-meat

meatData= data.table::fread("data/meat-supply-per-person.csv") 
colnames(meatData) = c("COUNTRY", "code", "year", "consumption")

unique(meatData$COUNTRY)

#check for missing data
meatData[,table(is.na(consumption))]
# no missing data
```


```{r merging all data, echo= FALSE, include= FALSE}
#selecting necessary info from meat data & only considering beef

#merging meat & anemia data
TotalData = merge(
  x = TotalAnemia,
  y = meatData,
  all.x = TRUE, all.y = FALSE
)

#download country code csv

countryCodes= data.table::fread("data/countryCodes.csv")
countryCodes = countryCodes %>%
  rename(CONTINENT = "Continent_Name", code= "Three_Letter_Country_Code") %>%
  select(code, CONTINENT)

#merging country code data frame & Total Data data frame so that the total data frame includes continent name
TotalData = merge(
  x = TotalData,
  y = countryCodes,
  all.x = TRUE, all.y = FALSE,
  by = "code"
)

TotalData = na.omit(TotalData)
```


```{r creating interactive plots, fig.height = 15, echo=FALSE, digits=2}


p3 = ggplot(TotalData, aes(x = year, y = (mildPct+modPct+sevPct)*100)) +
  geom_line(aes(color = COUNTRY)) + geom_smooth(method = "lm") + facet_wrap(~CONTINENT, nrow= 2)+
  labs(title = "Figure 1: Change in Total Anemia Prevalence in Each LMIC Over Time ")+
  xlab("Year") + ylab("Percentage of Women 15-49yo with Anemia")

p1 = ggplot(TotalData, aes(x = year, y = consumption)) +
  geom_line(aes(color = COUNTRY)) + geom_smooth(method = "lm") + facet_wrap(~CONTINENT, nrow= 2)+
  labs(title = "Figure 2: Change in Average Meat Consumption Per Capita in Each LMIC Over Time")+
  xlab("Year") + ylab("Average Meat Consumption (kg/capita)")


TotalDataAvg = TotalData[,.(
  consumption = mean(consumption,na.rm = TRUE),
  mildPct = mean(mildPct,na.rm = TRUE),
  modPct = mean(modPct,na.rm = TRUE),
  sevPct = mean(sevPct,na.rm = TRUE)
),by = "COUNTRY"]

library(ggpubr)
p4 = ggplot(TotalDataAvg,aes(x=consumption, y = (mildPct+modPct+sevPct)*100)) +
  geom_point(aes(color= COUNTRY)) +
  ggtitle("Figure 5: Anemia Prevalence vs Meat Consumption per Capita in LMICs")+
  ylab('Percentage of Women 15-49 with Anemia')+
  xlab('Meat Consumption (kg/capita/year)')+
  geom_smooth(method= "lm")


```

<br>

# Data Exploration

Multiple data sets were used for the purpose of exploring the question posed in the introduction. The data sets regarding the prevalence of anemia in 15-49 year old women in low and middle income countries (LMIC) were exported from the Global Health Data Exchange website (ghdx.healthdata.org) and was collected by the Institute of Health Metrics and Evaluation. The data sets regarding meat consumption per capita were exported from the Our World in Data site (ourworldindata.org) and was collected by the United Nations Food and Agricultural Organization (FAO). 

Following merging and clean up, the final data table included the consumption of meat per capita and the prevalence of mild, moderate, and severe anemia within women for 70 countries for each year between 2000 and 2017. Because both meat consumption and anemia prevalence are normalized to population size, it is possible to compare the two variables between countries with varying population sizes or across different years while population increases or decreases.


<br>

# Data Visualization



```{r echo=FALSE, fig.height = 15}
ggplotly(p3)
```
**Figure 1 ** This figure displays the trend of anemia prevalence in young women from 2000 to 2017. For ease of viewing, there are four subplots: one for each geographic region. Fortunately, there was an overall decrease in the prevalence of anemia in almost every LMIC, as shown by the best fit line in blue on each subplot. One clear and obvious exception to this downward trend is seen in Yemen, where the prevalence of anemia had increased by 10 percentage points over 17 years; this increase was likely due to the famine and instability present in this country. 



```{r echo=FALSE, fig.height = 15}
ggplotly(p1)
```
**Figure 2** This figure displays the periodic trend in average meat consumption within each country between 2000 and 2017 in units of kilograms of meat food per capita. In contrast to the trends in anemia prevalence, there was a slight increase in meat consumption in most countries.

<br>

```{r worldmap creation, echo=FALSE}

TotalData2017 = TotalData %>%
  filter(year == 2017) 

# light grey boundaries
l <- list(color = toRGB("grey"), width = 0.5)

# specify map projection/options
g <- list(
  showframe = FALSE,
  showcoastlines = TRUE,
  projection = list(type = 'Mercator')
)

fig <- plot_geo(TotalData2017)
fig <- fig %>% add_trace(
    z = ~(mildPct+modPct+sevPct)*100, color = ~(mildPct+modPct+sevPct)*100, colors = 'Blues',
    text = ~COUNTRY, locations = ~code, marker = list(line = l)
  )
fig <- fig %>% colorbar(title = '% of Women with Anemia')
fig <- fig %>% layout(
    title = 'Figure 3: Prevalence of Anemia in Women 15-49 in LMICs in 2017',
    geo = g
  )
figAnemia = fig

fig <- plot_geo(TotalData2017)
fig <- fig %>% add_trace(
    z = ~consumption, color = ~consumption, colors = 'Reds',
    text = ~COUNTRY, locations = ~code, marker = list(line = l)
  )
fig <- fig %>% colorbar(title = 'Meat Consumption (kg/capita)')
fig <- fig %>% layout(
    title = 'Figure 4: Meat Consumption per Capita in LMICs in 2017',
    geo = g
  )
figConsumption = fig

```


```{r echo=FALSE}
ggplotly(figAnemia)
```


```{r echo=FALSE}
ggplotly(figConsumption)
```
**Figures 3 & 4** The figures above visually display prevalence of anemia in young women and meat consumption per capita, respectively, in the Low and Middle Income Countries in 2017. The relationship between the two variables, when comparing the two figures, appears to be inversely proportional. It is interesting to note that in many south and southeastern Asian countries, meat consumption is low due to meatless diets common in the cultures and religion found there.


```{r echo=FALSE}
ggplotly(p4)
```


**Figure 5** Figure 5 takes the average anemia prevalence and meat consumption over the time period of 2000 to 2017 to compare between the two variables. There is a negative linear relationship between meat consumption and anemia prevalence in LMICs, as shown by the slope of the linear best fit line. As meat consumption increases in a country, the prevalence of anemia decreases. Most African countries have low values for meat consumption per capita and high prevalence of anemia. Many South American countries have higher values for meat consumption and a lower prevalence of anemia. There are not many points which fall directly on the best fit line or in its immediate vicinity, indicating that the correlation might not be that large or that there are many outliers.

# Summary


Based on the data presented above, there is a high negative correlation between meat consumption and anemia prevalence. There are likely many other factors affecting the high levels of anemia found in these LMICs, some of which are also related to diet. This is only a preliminary exploration of the available data and further research could need to be done to consider the effect of other factors, such as the consumption of other minerals and vitamins or the prevalence of  blood and digestive-related diseases. 

Furthermore, it is positive to see an overall decrease in the prevalence of anemia in almost all countries. This could be indicative of improving general female health in LMICs as well.

The full report, which includes more in-depth methods & analysis, can be viewed and downloaded: 